Chasing \rainbows{} and ocean glint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-01 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight that is reflected by an exoplanet can reveal angular features that are due to ocean glint, and to rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \hwo's coronagraph may obscure the phase angles where such characteristic features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
Planets that orbit close to edge-on have accessible scattering angles limited by coronagraph obscuration. 
%
We use \hwo'~catalog of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering by gas. 
%
We find that any exo-Earths in systems in this catalog will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The glint signature at planetary scattering angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to \hwo~in $\sim\num{14}$ systems, and the rainbow due to water clouds at phase angles of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} \IWA{} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with the planet's orbital eccentricity and scales approximately inversely with the \IWA{}.  \end{abstract}

@@

%--------------------------------------------------------------------------
% Figure 4
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
% Figure 5
%--------------------------------------------------------------------------
\begin{figure}
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/orb-grid.pdf}
   \caption{
        The phase angles accessible to a direct imaging mission 
        can be limited by either the orbital inclination angle $i$ or the coronagraphic \IWA{}.
        %
        If the orbit is close to face-on (top panels), the phase angles probed along the orbit are limited by the inclination angle $i$; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm i$. 
        %
        For orbits close to edge-on (bottom panels) the planet attains all 
        phases, but the phase angles near \qty{0}{\degree} and \qty{180}{\degree} are not observable due to obscuration by the coronagraph; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
    \script{plot-orb-grid.py}
\end{figure}
%--------------------------------------------------------------------------

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight that is reflected by an exoplanet can reveal angular features that are due to oceans, and are sensitive to rainbows and other phenomena that are due to scattering by cloud particles. 
%
Direct imaging missions are optimised for planets near quadrature, and \hwo's coronagraph may obscure the phase angles where such characteristic features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
Planets in close-to-edge-on orbits have accessible phase angles limited by coronagraph obscuration. 
%
We use \hwo'~catalog of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering by gas. 
%
We find that any exo-Earths in systems in this catalog will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The glint signature at planetary scattering angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to \hwo~in $\sim\num{14}$ systems, and the rainbow due to water clouds at phase angles of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} \IWA{} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with the planet's orbital eccentricity and scales approximately inversely with the \IWA{}.  \end{abstract}

@@

%--------------------------------------------------------------------------
% Figure 4
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-figure-4-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
% Figure 5
%--------------------------------------------------------------------------
\begin{figure}
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/figure-5-orb-grid.pdf}
   \script{create-figure-5-orb-grid.py}
   \caption{
        The phase angles accessible to a direct imaging mission 
        can be limited by either the orbital inclination angle $i$ or the coronagraphic \IWA{}.
        %
        If the orbit is close to face-on (top panels), the phase angles probed along the orbit are limited by the inclination angle $i$; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm i$. 
        %
        For orbits close to edge-on (bottom panels) the planet attains all 
        phases, but the phase angles near \qty{0}{\degree} and \qty{180}{\degree} are not observable due to obscuration by the coronagraph; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
\end{figure}
%--------------------------------------------------------------------------
